A weakly structured stem for human origins in Africa

### 2022-01-30 paper.tex
Addressing comments, Results

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}. These
approaches are more challenging to interpret and fit to data. However, new
population genetic inference tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with modern humans in Eurasia, several recent articles
have shown that introducing an archaic ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific archaic remains, and the possibility of
finding ancient DNA evidence (e.g., \citet{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic admixture model, leaving out other plausible models
(Figure~\ref{fig:supp-possible-models} and \citet{Henn2018-rf}). Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African ancestry. While ancient DNA from
Eurasia has helped us understand early human history outside of Africa, there
is no comparably ancient DNA to elucidate early history in Africa.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
included the British (GBR) from the 1000 Genomes Project \sgcomment{phase 3} in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for archaic gene flow from Neanderthals in non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550kya from a common
stem. For each population, we computed low-order allele frequency and linkage
disequilibrium (LD) statistics that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

@@

Regardless of the model choice for early epochs, inference of human demographic
history in the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. We conclude that, regardless of model
specification, geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
\sgcomment{Alt formulation: Thus all models suggest that broad geographic patterns of contemporary population structure date back to MIS 5}
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (Supp.
Methods). The time of divergence of Western and Eastern African populations was
set to 60kya, just prior to the split of Eurasians and East Africans 50kya. All
models predict relatively high subsequent gene flow between the two regions
($m\approx2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
\sgcomment{Here it is not clear whether we are still talking about parameters that we fixed. I think we are also missing a clear description of the best-fit model. That *could* go after the description of the model uncertainty}
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE}
\citep{Alexander2009-sw} (Figure~\ref{fig:1}).

@@

To account for population structure prior to 135kya, we allowed \sgcomment {alt: three of our four models allowed} for two or more
“stem” populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or “merger”) events from the
other stem population, so that diversifying branches are formed through the
\sgcomment{This sentence is not clear to me. Do we mean: In other words, the more recent populations are formed by mergers of multiple ancestral populations }
coalescence of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic introgression and
fragmentation-and-coalescence models.

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger scenario and $LL \approx -115,500$ vs. $-126,600$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this roughly compares
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
the early branches remain closely related (Figure 3).

Bootstrap-based confidence intervals are relatively narrow within each
specified model
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. As expected, model assumptions
have a larger impact on parameter estimates (and thus, uncertainty). To convey model uncertainty, we
highlight features of two inferred models with high likelihoods, the
“multiple-merger” and the the “continuous-migration” model that both allow
migration between stem branches. These differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

\sgcomment{I don't love this new structure because it breaks the model description in two parts separated by the bootstrap. Maybe we have to, but I would have preferred a unified description of the models. 
For example, we could start this section by the bootstrap discussion ("In all models, likelihood differences are large and bootstrap confidence interval narrow within each specified model ...  We therefore highlight two models".
Then we describe model 1, describe model 2, and contrast the two. 
 ) }
Under the continuous-migration model, one of the two stems diverges into
lineages leading to contemporary populations in western, southern and eastern
Africa, and the other (Stem 2) contributes variable ancestry to those
populations. This migration from Stem 2 is highest with the Mende
($m=1.6\times10^{-4}$) compared to the Nama and East African populations
($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
allowed to occur until the relatively recently demographic events 5kya. A
sampled lineage from the Nama, Mende, and Gumuz have probabilities of being in
Stem 2 at the time of Stem 1 expansion (135ka) of approximately $0.145$, $0.2$,
and $0.13$, respectively, though these probabilities change over time,
precluding the notion of a fixed admixture proportion.

>>>>

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}. These
approaches are more challenging to interpret and fit to data. However, new
population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with modern humans in Eurasia, several recent articles
have shown that introducing an archaic ghost population contributing to African
populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models
\citep{Plagnol2006-lt,Hammer2011-bx,Hsieh2016-gk,Hey2018-pw,Ragsdale2019-nt,Durvasula2020-td,Lorente-Galdos2019-vz,Durvasula2020-td}.
This has driven speculation about the geographic range of this ghost
population, possible links to specific archaic remains, and the possibility of
finding ancient DNA evidence (e.g., \citet{Hsieh2016-gk}). However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic admixture model, leaving out other plausible models
(Figure~\ref{fig:supp-possible-models} and \citet{Henn2018-rf}). Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African ancestry. While ancient DNA from
Eurasia has helped us understand early human history outside of Africa, there
is no comparably ancient DNA to elucidate early history in Africa.

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierre Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendents of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2019-wd}), and Eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2019-wd} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:1}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for archaic gene flow from Neanderthals in non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged 550kya from a common
stem. For each population, we computed low-order allele frequency and linkage
disequilibrium (LD) statistics that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

@@

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. We conclude that, regardless of model
specification, geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
\sgcomment{Alt formulation: Thus all models suggest that broad geographic patterns of contemporary population structure date back to MIS 5}
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

Given this consistency in inferred recent history and the numerical challenge
of optimizing a large number of parameters, we fixed several parameters related
to recent population history so as to focus on more ancient events (Supp.
Methods). The time of divergence of Western and Eastern African populations was
set to 60kya, just prior to the split of Eurasians and East Africans 50kya. All
models infer relatively high subsequent gene flow between the two regions
($m\approx2\times10^{-4}$). Admixture from Neanderthals to Europeans directly
following the out-of-Africa migration was set to 1.5\% at 45kya (Supp.
Methods). Back-to-Africa gene flow at the beginning of the Holocene primarily
\sgcomment{Here it is not clear whether we are still talking about parameters that we fixed. I think we are also missing a clear description of the best-fit model. That *could* go after the description of the model uncertainty}
affected the ancestors of the Ethiopian agricultural populations, comprising
over half of their genetic ancestry, estimated to be 64--65\%. The past 5,000
years also saw major demographic changes, including strong population growth
for Western Africans as they specialized in yam and oil palm agriculture
(estimated 3-fold growth). We observe significant gene flow from the Amhara and
Oromo into the Nama, a signal which is likely a proxy for the movement of
Eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions inferred by \texttt{ADMIXTURE}
\citep{Alexander2009-sw} (Figure~\ref{fig:1}).

@@

To account for population structure prior to 135kya, three of our four models allowed for two or more
“stem” populations which could diverge either before or after the Neanderthal
split. We considered models both with and without migration between these stem
populations, and in both cases we tested two different types of gene exchange
during the expansion phase. 1) One of the stem population expands (splits into
contemporary populations), and the other stem population(s) has continuous
symmetric migration with those populations; or 2) one or more of the stem
populations expands, with instantaneous pulse (or “merger”) events from the
other stem population, so that recent populations are formed by mergers of multiple ancestral populations. Depending on parameter values,
this scenario encompasses archaic introgression and
fragmentation-and-coalescence models. For many parameters, confidence intervals based on bootstrapping are relatively narrow
(Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}),
reflecting an informative statistical approach. Model assumptions
have, however, a larger impact on parameter estimates (and thus, uncertainty). To convey model uncertainty, we
highlight features of the two inferred models with high likelihoods. These are referred to as the
“multiple-merger” and the the “continuous-migration” model. Both allow for
migration between stem branches, but differ primarily in the timing of the
early divergence of stem populations and their relative $N_e$
(Figure~\ref{fig:2}). The two models also differ in the mode of divergence
during the Middle Stone Age.

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration ($LL \approx -102,600$ vs.
$-107,700$ in the merger model and $LL \approx -115,500$ vs. $-126,600$ with
continuous migration). With continuous migration between stems, population
structure extends back to 1.1--1.4Mya (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.43\times10^{-5}-1.16\times10^{-4}$. For comparison, this is similar
to inferred migration rates between connected contemporary populations over the
past 50kya (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic admixture models (Figure~\ref{fig:supp-possible-models}C) as
the early branches remain closely related (Figure 3).

\sgcomment{I don't love this new structure because it breaks the model description in two parts separated by the bootstrap. Maybe we have to, but I would have preferred a unified description of the models. 
For example, we could start this section by the bootstrap discussion ("In all models, likelihood differences are large and bootstrap confidence interval narrow within each specified model ...  We therefore highlight two models".
Then we describe model 1, describe model 2, and contrast the two. 
 )} \bmh{what do you think of the revised text?}

Under the continuous-migration model, one of the two stems diverges into
lineages leading to contemporary populations in western, southern and eastern
Africa, and the other (Stem 2) contributes variable ancestry to those
populations. This migration from Stem 2 is highest with the Mende
($m=1.6\times10^{-4}$) compared to the Nama and East African populations
($m=5.8\times10^{-5}$ and $3.1\times10^{-5}$, respectively), with migration
allowed to occur until the relatively recently demographic events 5kya. A
sampled lineage from the Nama, Mende, and Gumuz have probabilities of being in
Stem 2 at the time of Stem 1 expansion (135ka) of approximately $0.145$, $0.2$,
and $0.13$, respectively, though these probabilities change over time,
precluding the notion of a fixed admixture proportion.
